---
title: "Figure 1: short-term GPP–wood-growth coupling"
author: "Ngoc Nguyen"
date: "2025-10-27"
output:
  html_document:
    toc: true
    toc_depth: 2
---

This document reproduces Figure 1 from annual TRENDYv10 outputs and the
site-level observations compiled by Cabon et al. (2022).

## Run instructions

Run from the repository's `code/` directory:

```bash
Rscript -e 'rmarkdown::render("Figure1_code_clean.Rmd")'
```

Alternatively, set `SOURCE_SINK_ROOT` to the repository root before rendering:

```bash
SOURCE_SINK_ROOT=/path/to/source_sink_dynamics \
  Rscript -e 'rmarkdown::render("code/Figure1_code_clean.Rmd")'
```

Expected inputs and generated outputs are defined in the setup chunk below.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(dplyr)
library(ggplot2)
library(patchwork)
library(purrr)

# Set SOURCE_SINK_ROOT when rendering from outside code/.
project_root <- normalizePath(
  Sys.getenv("SOURCE_SINK_ROOT", unset = ".."),
  mustWork = TRUE
)

data_dir <- normalizePath(
  Sys.getenv(
    "CABON_DATA_DIR",
    unset = file.path(project_root, "data", "cabon")
  ),
  mustWork = TRUE
)
trendy_dir <- file.path(project_root, "results", "TRENDYv10", "31_site_weighted")
derived_dir <- file.path(project_root, "results", "datasets")
figure_dir <- file.path(project_root, "results", "figures")

dir.create(derived_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

models <- c(
  "ISBA-CTRIP", "CABLE-POP", "CLM5.0", "LPJ-GUESS", "LPX-Bern",
  "ORCHIDEE", "ORCHIDEEv3", "CLASSIC-N", "CLASSIC"
)
scenarios <- c("S0", "S1", "S2")
wood_variable <- "cWood"

required_inputs <- c(
  file.path(data_dir, "Cabonetal_site_info.csv"),
  file.path(data_dir, "rw_onsite.csv"),
  file.path(data_dir, "flux_onsite.csv")
)

missing_inputs <- required_inputs[!file.exists(required_inputs)]
if (length(missing_inputs) > 0) {
  stop("Missing required input files:\n", paste("-", missing_inputs, collapse = "\n"))
}
```

## Reusable functions

```{r functions}
assert_files_exist <- function(paths, context) {
  missing <- paths[!file.exists(paths)]
  if (length(missing) > 0) {
    stop(context, ":\n", paste("-", missing, collapse = "\n"))
  }
}

detrended_correlation <- function(x, y, min_n = 4L) {
  values <- data.frame(x = x, y = y) |>
    filter(is.finite(x), is.finite(y)) |>
    mutate(time_index = row_number())

  if (nrow(values) < min_n || sd(values$x) == 0 || sd(values$y) == 0) {
    return(tibble(n = nrow(values), correlation = NA_real_, p_value = NA_real_))
  }

  x_residual <- residuals(lm(x ~ time_index, data = values))
  y_residual <- residuals(lm(y ~ time_index, data = values))
  test <- cor.test(x_residual, y_residual, method = "pearson")

  tibble(
    n = nrow(values),
    correlation = unname(test$estimate),
    p_value = test$p.value
  )
}

model_input_paths <- function(model, scenario) {
  model_dir <- file.path(trendy_dir, model)
  c(
    gpp = file.path(model_dir, sprintf("gpp_%s_31_site_weighted_yearly_mean.csv", scenario)),
    wood = file.path(model_dir, sprintf("%s_%s_31_site_weighted_yearly_mean.csv", wood_variable, scenario)),
    sites = file.path(model_dir, sprintf("%s_site_info.csv", model))
  )
}

model_statistics_path <- function(model, scenario) {
  file.path(
    derived_dir,
    sprintf(
      "%s_%s_detrended_r_normal_delta_31_gpp_TRENDYv10_temporalMatch_not_scale%s.csv",
      model, wood_variable, scenario
    )
  )
}

significance_label <- function(p_value) {
  case_when(
    is.na(p_value) ~ "",
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**",
    p_value < 0.05 ~ "*",
    TRUE ~ ""
  )
}

safe_paired_test <- function(model_values, observed_values) {
  keep <- is.finite(model_values) & is.finite(observed_values)
  if (sum(keep) < 2) return(NA_real_)
  t.test(model_values[keep], observed_values[keep], paired = TRUE)$p.value
}
```

## Site metadata

```{r site-metadata}
site_metadata <- read.csv(file.path(data_dir, "Cabonetal_site_info.csv")) |>
  filter(On.site.RW) |>
  transmute(SITE_ID = Site, Start, End)

if (anyDuplicated(site_metadata$SITE_ID)) {
  stop("Cabon site metadata contains duplicate SITE_ID values.")
}
```

## Calculate model correlations

For every model and scenario, this section aligns GPP and wood carbon by year
and grid cell, limits each site to its observational period, linearly detrends
both series, and calculates their Pearson correlation.

```{r model-correlations}
calculate_model_statistics <- function(model, scenario) {
  paths <- model_input_paths(model, scenario)
  assert_files_exist(paths, sprintf("Missing inputs for %s %s", model, scenario))

  gpp <- read.csv(paths[["gpp"]]) |>
    filter(is.finite(gpp))

  wood <- read.csv(paths[["wood"]]) |>
    filter(is.finite(.data[[wood_variable]]), .data[[wood_variable]] != 0)

  # CLM5.0 output ends before 2020 in this analysis.
  if (model == "CLM5.0") {
    gpp <- filter(gpp, year < 2020)
    wood <- filter(wood, year < 2020)
  }

  site_coordinates <- read.csv(paths[["sites"]])

  aligned <- inner_join(gpp, wood, by = c("year", "lat", "lon")) |>
    inner_join(site_coordinates, by = c("lat", "lon")) |>
    inner_join(site_metadata, by = "SITE_ID") |>
    filter(year >= Start, year <= End) |>
    arrange(SITE_ID, year) |>
    mutate(gpp = gpp * 1000)

  statistics <- aligned |>
    group_by(lat, lon, SITE_ID) |>
    group_modify(~ detrended_correlation(.x[[wood_variable]], .x$gpp)) |>
    ungroup() |>
    rename(
      r_detrend_delta = correlation,
      p_detrend_delta = p_value
    ) |>
    filter(is.finite(r_detrend_delta))

  write.csv(
    statistics,
    model_statistics_path(model, scenario),
    row.names = FALSE
  )

  statistics
}

model_statistics <- tidyr::crossing(model = models, scenario = scenarios) |>
  mutate(statistics = map2(model, scenario, calculate_model_statistics))
```

## Calculate observational correlations

```{r observation-correlations}
ring_width <- read.csv(file.path(data_dir, "rw_onsite.csv"))

annual_gpp <- read.csv(file.path(data_dir, "flux_onsite.csv")) |>
  filter(QC > 0.7) |>
  summarise(GPP_dtrd = mean(GPP_dtrd, na.rm = TRUE), .by = c(Year, Site))

observations <- inner_join(ring_width, annual_gpp, by = c("Year", "Site")) |>
  filter(is.finite(GPP_dtrd), is.finite(RWI_dtrd)) |>
  group_by(IDcrn) |>
  group_modify(~ detrended_correlation(.x$GPP_dtrd, .x$RWI_dtrd)) |>
  ungroup() |>
  mutate(SITE_ID = substr(IDcrn, 1, 6)) |>
  summarise(
    r_observed = mean(correlation, na.rm = TRUE),
    p_observed = mean(p_value, na.rm = TRUE),
    .by = SITE_ID
  )
```

## Assemble matched comparisons

Model–observation paired tests are explicitly aligned by `SITE_ID` before the
test is run.

```{r matched-comparisons}
model_results <- model_statistics |>
  select(model, scenario, statistics) |>
  tidyr::unnest(statistics)

matched_results <- model_results |>
  inner_join(observations, by = "SITE_ID")

paired_tests <- matched_results |>
  summarise(
    p_value = safe_paired_test(r_detrend_delta, r_observed),
    .by = c(model, scenario)
  ) |>
  mutate(asterisk = significance_label(p_value))

plot_data <- matched_results |>
  select(model, scenario, SITE_ID, r_detrend_delta) |>
  bind_rows(
    matched_results |>
      distinct(model, SITE_ID, r_observed) |>
      transmute(
        model,
        scenario = "obs",
        SITE_ID,
        r_detrend_delta = r_observed
      )
  ) |>
  mutate(scenario = factor(scenario, levels = c("obs", scenarios)))
```

## Generate Figure 1

```{r figure-1, fig.width=7, fig.height=6.5}
make_model_panel <- function(model_name) {
  panel_data <- filter(plot_data, model == model_name)
  n_sites <- n_distinct(panel_data$SITE_ID)

  annotations <- panel_data |>
    summarise(y_position = min(1.1, max(r_detrend_delta, na.rm = TRUE) + 0.05), .by = scenario) |>
    left_join(
      paired_tests |>
        filter(model == model_name) |>
        select(scenario, asterisk),
      by = "scenario"
    ) |>
    mutate(asterisk = coalesce(asterisk, ""))

  ggplot(panel_data, aes(scenario, r_detrend_delta)) +
    geom_boxplot(
      linewidth = 0.3,
      alpha = 0.5,
      outlier.shape = NA,
      colour = "#3c5488",
      fill = "#3c5488"
    ) +
    geom_jitter(
      width = 0.15,
      size = 0.5,
      alpha = 0.5,
      colour = "#3c5488"
    ) +
    geom_text(
      data = annotations,
      aes(scenario, y_position, label = asterisk),
      inherit.aes = FALSE,
      size = 3.5
    ) +
    annotate(
      "text",
      x = 0.6,
      y = -0.8,
      label = sprintf("%s (n = %d)", model_name, n_sites),
      hjust = 0,
      vjust = 1,
      size = 3.5
    ) +
    coord_cartesian(ylim = c(-1, 1.2)) +
    labs(
      x = NULL,
      y = expression(r[Delta * italic(cWood) * "," ~ italic(GPP)])
    ) +
    theme_bw(base_size = 12) +
    theme(
      legend.position = "none",
      panel.grid = element_blank()
    )
}

plots <- map(models, make_model_panel)

for (index in seq_along(plots)) {
  if (index %% 3 != 1) {
    plots[[index]] <- plots[[index]] +
      theme(axis.title.y = element_blank(), axis.text.y = element_blank())
  }
  if (index <= 6) {
    plots[[index]] <- plots[[index]] +
      theme(axis.text.x = element_blank())
  }
}

figure_1 <- wrap_plots(plots, ncol = 3) +
  plot_annotation(tag_levels = "a")

ggsave(
  filename = file.path(figure_dir, "Figure1.pdf"),
  plot = figure_1,
  width = 7,
  height = 6.5,
  units = "in"
)

figure_1
```

## Session information

```{r session-information}
sessionInfo()
```
